---
title: Descriptive statistics in R, part II
author: J. Winget
date: '2018-04-23'
slug: descriptives-part-ii
categories: []
tags:
  - statistics
  - R
header:
  caption: ''
  image: ''
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE)
```

Moving on from frequencies and tables, which were covered in [part I](https://jeremyrwinget.netlify.com/post/descriptives-part-i/), let's now focus on other ways to summarize our data (e.g., mean, standard deviation). There are a lot of ways to divide a topic like descriptive statistics, and R can further complicate this seemingly simple task. It's been said R does a great job of making complex procedures simple, but it also has the tendency to make simple tasks complex. Because there are many methods of calculating descriptive statistics in R, and because some of these methods are more applicable in certain situations over others, there is a lot of material covered in this post. To keep the post length manageable, part II will cover some ways to calculate descriptive statistics in base R and some common functions in popular packages. In part III, we'll focus on some other common approaches for obtaining descriptive statistics, and I'll finish by discussing my thoughts on the packages between parts II and III overall.

For this post and the next, I will be using the ```gapminder``` dataset in the ```gapminder``` package. The dataframe contains data on life expectancy, GDP per capita, and population by country, continent, and year. It also contains 1704 observations and 6 variables. The data come from [Gapminder](https://www.gapminder.org), which is an independent Swedish foundation with no political, religious, or economic affiliation.

```{r}
library(gapminder)
str(gapminder)
```

Again, if you're familiar with SPSS, you can get means, SDs, and pretty much all of your general descriptive statistics through either the FREQUENCIES, (CON)DESCRIPTIVE, or EXAMINE commands. There are some functions in R that are pretty similar to the output SPSS displays, but there are also a few cool functions that are exclusive to R. 

#Options in base R

R's built-in function for univariate statistics is ```summary```. 

```{r}
summary(gapminder)
```

This output is pretty sparse. It lacks percentages for factors (i.e., country and continent), frequencies and percentages for numeric variables (i.e., lifeExp and gdpPercap), number of nonmissing values, and so on. However, it is very compact. This compactness allows ```summary``` to work with a wide range of object whereas other functions (e.g., ```describe``` described below) only work with dataframes, vectors, matrices, or formulas. The numbers labeled "1st Qu." and "3rd Qu." are the first and third quartiles, or the 25th and 75th percentiles, respectively. 

If the dataset contains any variable labels, those will be ignored. It also doesn't show the overall count, but you could manually compute this by summing up the counts in the categorical variables. Then again, what's the point of using a powerful program like R if you just end up manually computing things?! ```summary``` also doesn't display standard deviation, and it's not tidy.

If you need to summarize data by group, which often needs to be done, you can do this with base R as well. However, the output isn't nicely formatted, which can make it difficult to compare statistics across groups, especially if there are a large number of categories (like there are in country variable). Since there are so many categories in the below example, I wrapped the command in the ```head``` function. This allows me to conserve space by only displaying the first six cases of the output.

```{r}
head(by(gapminder, gapminder$country, summary))
```

These functions are certainly useful, but there are other more informative and flexible options for summarizing our data outside of base R.

#describe, from the Hmisc package

The ```Hmisc``` package offers a wide selection of functions that have a more comprehensive output than the base R functions. One of these is the ```describe``` function. It's similar to the ```summary``` described above, but one nice feature of the ```describe``` function is that it provides frequencies on nonfactors as well as factors, as long as they do not have too many values.

```{r}
library(Hmisc)
describe(gapminder)
```

Unlike SPSS, the ```describe``` function does not provide percentages that include missing values. You can change that by setting the ```exclude.missing``` argument to ```FALSE```. ```describe``` will automatically provide a table of frequencies whenever a variable has no more than 20 unique values. Beyond that, it will report the five largest and five smallest values, just like SPSS's EXPLORE procedure. Also, for some items (not pictured in this example), the ```describe``` function will display the variable prompts themselves (e.g., the survey question participants responded to) in the output. This can be accomplished by adding variable labels (e.g., through the ```label``` function in the ```Hmisc``` package).

SPSS users may find the output user-friendly, but it isn't tidy and can be hard to work with at times. There is also no summary by group function, but you could get around this by wrapping ```describe``` in the ```by()``` function mentioned earlier. One downside of this approach, though, is that the output becomes lengthy and comparing statistics between groups becomes challenging.

#describe and describeBy, from the psych package

Not to be confused with ```describe``` from the ```Hmisc``` package, the ```describe``` function from the ```psych``` package has a strikingly different output!

```{r}
library(psych)
describe(gapminder)
```

This output contains all six of the variables in the dataset; however, the categorical variables (i.e., country and continent) display summary stats that you'd typically associate with numeric fields. This is because the function's default is to recode categories as numbers. When the function recodes a categorical variable into a numeric variable, it denotes this by adding an `*` at the end of the variable name. **Thus, statistics for variables marked with `*` should be interpreted cautiously (if at all).**

If you're dealing with categorical data that you *do not* want converted into numerical data, ```psych::describe``` may not be the best function to use. However, for genuinely numeric data though, it provides most of the key statistics and a few bonus stats (e.g., skew, kurtosis, SE). However, it does not indicate where data are missing. The output is fairly tidy except for it using rownames to represent the variable the statistics are summarizing, but this is a pretty common feature across packages.

Another nice feature about the ```psych``` package is that it has a specific summary by group function: ```describeBy```. If we wanted the statistics for each continent in the ```gapminder``` dataset, we would simply use the following code:

```{r}
describeBy(gapminder, gapminder$continent)
```

All the essential statistics are there! This output more condensed than output from the ```by()``` function, but it's still difficult to compare the same statistic across groups. It's also not tidy. However, the ```mat``` argument let's you produce a matrix from the above output.

```{r}
describeBy(gapminder, gapminder$continent, mat = TRUE)
```

It's not pretty, but it does have the added benefit of producing a potentially useful dataframe that you could tidy up.

#descr and dfsummary, from the summarytools package

Stating with ```descr```, the first thing to note is that it (like ```psych::describe```) only works with numerical data. However, unlike ```psych::describe```'s asterisk, we're altered by an actual warning message displayed at the top of the output showing which columns have been ignored. There's also a record count and a nice variety of summary statistics. The output also displays some information on missing data (e.g., ```Pct.Valid``` is the proportion of data that *is not* missing). 

```{r}
library(summarytools)
descr(gapminder)
```

There's a ```transpose``` argument for those who prefer to arrange their variables by row and summary statistics as columns.

```{r}
descr(gapminder, transpose = TRUE)
```

```summarytools``` also includes a more comprehensive summarizing function called ```dfSummary``` that is intended to summarize a whole dataframe.

```{r}
dfSummary(gapminder)
```

The output from ```dfSummary``` is much more detailed than many of the functions we've looked at so far, and it can deal with both categorical and numeric variables. It displays commonly used summary statistics, information on missing data, sample sizes, and variable type. It also uses a textgraph to show rough distributions (but these pale in comparison to the ```skimr``` package described below). However, one limitation is that there is no special grouping feature.

#skim, from the skimr package

The ```skim``` function summarizes the record and variable count. It is fantastic for handling both categorical and numeric data, and displays accurate summary stats for each. For readability, it separates variables into different sections dependent on data type, which in turn makes for quick interpretation given that different summary stats are relevant for different data types. It also clearly reports missing data, and has all the most common summary stats. However, it doesn't display some of the bonus content that ```psych::describe``` does such as SE and range. 

```{r}
library(skimr)
skim(gapminder)
```

```skim```'s output also works very well with ```group_by``` from the ```dplyr``` package, which means you can easily produce summaries by group:

```{r}
library(dplyr)
group_by(gapminder, continent) %>% 
  skim()
```

Because the output is arranged by the grouping variable (i.e., continent in the current example), sometimes it can be difficult to compare summary statistics across groups. However, this method is probably one of the easier/nicer ways to view these results at a glimpse. However, with a little more manipulation, it's actually pretty easy to directly compare statistics across groups. Despite the above ```skim``` output not looking tidy, there is a tidy dataframe-like representation of each combination of variable and statistic happening behind the scenes. For example:

```{r}
data <- group_by(gapminder, continent) %>% 
          skim()
head(data)
```

It's not exactly human-reader friendly, but after some deciphering, you might be able to see it produces a long table that contains one row for every combination of group, variable, and summary statistic. This means you can use standard methods of dataframe manipulation to process your summary statistics.

For instance, using the ```tidyverse```, we can examine the mean and SD of the lifeExp variable and compare the results between each continent in the ```gapminder``` dataset.

```{r}
data %>% 
  filter(variable == "lifeExp", stat %in% c("mean", "sd")) %>% 
  ggplot(aes(x = continent, y = value)) +
  facet_grid(stat ~ ., scales = "free") +
  geom_col()
```

Overall, ```skimr``` is a simple to use summary function that can be used with pipes and displays. You can modify the default summary statistics, as can the default formatting. Support for data frames and vectors is included, and you can implement your own skim methods for specific object types. So, I encourage you to check out the [skimr documentation](https://cran.r-project.org/web/packages/skimr/skimr.pdf) for further details on these topics.

#Summary

That about wraps it up for part II. In this post, I covered five packages for summarizing your data in R.

1. ```summary``` and ```by``` in base R
2. ```describe``` in the ```Hmisc``` package
3. ```describe``` in the ```psych``` package
4. ```descr``` and ```dfSummary``` in the ```summarytools``` package
5. ```skim``` in the ```skimr``` package

I tend to use ```skimr``` as my go-to package for descriptive statistics, mostly because I can seamlessly integrate it with my tidy workflow. I used ```psych::describe``` as my main data summary function for a very long time because it provided me with descriptive statistics that many other functions neglect. However, since ```skimr``` allows you to modify the default summary statistics it displays, I find myself using it more and more. That being said, there are many times I'll use base R's ```summary``` function because of its simplicity and flexibility with other functions. 

However, these five packages are just the tip of R's descriptive statistic iceberg. Stay tuned for more descriptive statistics in R part III, coming soon!
